Ace: An Actively-Typed Language and Compilation Environment for High-Performance Computing

### 2013-04-25 ace/ace.tex
high-level design

In this paper, we introduce Ace, a programming language targeting professional end-users as well as researchers across high-performance computing and related domains. Ace has followed a principled design methodology guided by this more comprehensive set of design criteria in order to avoid many of the issues that have hindered previous language-based approaches in high-performance computing. These criteria, this design methodology, and the novel mechanisms and designs developed to satisfy these criteria constitute the generalizable contributions of this paper. We also hope that Ace itself will be useful to the HPC community\footnote{\small Ace is openly available at http://acelang.org/.}, and we describe several use cases and an initial case study as a means to preliminarily validate it's utility.

@@

The examples in Section \ref{usage} are based on an internalization of the OpenCL kernel programming language as a library. This library is built atop a novel extensibility mechanism that we call \emph{active typechecking and translation} (AT\&T) and detail in Section \ref{att}. This mechanism relies on the core idea of representing types as metalanguage objects. We refer to such types as \emph{active types} by analogy to \emph{active libraries} \cite{activelibraries} (see Section \ref{related}). Unlike prior approaches to extensibility that take a language-global view, which impacts interoperability, the AT\&T mechanism features a guarantee that extensions are composable by construction.

%Despite its simplicity and safety, this mechanism is flexible enough to define the entirety of the OpenCL kernel language (which is derived directly from C99) and, combined with Ace's type propagation scheme, substantially enhance it's ease-of-use. 
To demonstrate the flexibility of this mechanism beyond OpenCL, we continue in Section \ref{usecases} by outlining examples of other, higher-level parallel abstractions that can be cleanly implemented using AT\&T, including global address spaces, message passing and functional data parallelism. First-class support for each of these has required a new language in the past. AT\&T allows compile-time logic to be safely included within libraries, and thus these first-class abstractions can coexist within a single program or workflow.

Ace can be used as a standalone language via the \verb|acec| compiler and also as an interactive compilation environment. This mode of use, described in Section \ref{compenv}, uses just-in-time specialization to deeply integrate the widely-adopted \verb|numpy| library (including its internal type system) and OpenCL's host API (by a mechanism also available to other host APIs, such as CUDA's) with the Ace compiler itself to enable the direct invocation of Ace functions from within Python scripts with minimal syntactic overhead. Achieving this integration also makes use of Ace's representation of types.

To demonstrate the utility of Ace for scientific workloads, we describe in Section \ref{casestudy} a case study where Ace was used as the underlying compilation environment for a scientific simulation framework. This framework was used to  specify and efficiently execute thousands of realizations of a large stochastic neurobiological circuit model on clusters of GPUs.

We conclude in Sections \ref{related} and \ref{discussion} with related work and a discussion of the limitations of Ace at the time of writing, as well as a discussion of planned future work building upon the concept of active typing.

@@

\section{Design and Usage}\label{usage}
It is now widely-acknowledged that static type systems are fundamental tools for performance and verifiability, both critical criteria in HPC, so Ace is fundamentally a statically-typed language. 
To achieve continuity, Ace is implemented as a library in the ubiquitous scripting language Python. The Ace grammar is precisely that of Python, and it share's Python's packaging system, so any tools that work with Python files and packages can immediately be used with Ace.

Another important tool that has become increasingly prominent is the use of code generation techniques to optimize code and orchestrate scientific workflows, so Ace includes a powerful

\begin{itemize}
\item Types are fundamental tools for both performance and verifiability.
\item Programmers in high-performance computing tend to be wary of abstraction mechanisms that necessitate run-time overhead.
\item Compile-time code generation is often used to optimize code and orchestrate scientific workflows.
\item Scripting languages, such as Python, are particularly appreciated for their ease-of-use.
\item Library-based abstractions are more highly-adopted, but language-integrated abstractions are easier to use \cite{xxx}.
\end{itemize}

\begin{codelisting}
\lstinputlisting{hello.py}
\caption{\texttt{[hello.py]} A basic Ace program demonstrating the two-phase structure of Ace programs and libraries.}
\label{hello}
\end{codelisting}
\begin{codelisting}
\begin{lstlisting}[style=Bash]
$ `acec hello.py`
Hello, compile-time world!
Goodbye, compile-time world!
$ `cat hello.cl`
__kernel void main() {
    printf("Hello, run-time world.");
}\end{lstlisting}
\caption{Compiling \texttt{hello.py} using the \texttt{acec} compiler.}
\label{helloout}
\end{codelisting}

\section{Structure and Usage}
%Ace is an {\em actively-typed, language-integrated compilation environment}. Let us begin by defining each of these terms. 
%A {\em compilation environment} is a system that allows developers to programmatically control aspects of compilation directly. The language used to do so is called the {\em metalanguage} while the grammar that programs themselves are written in is called the {\em source grammar}. The metalanguage of Ace is Python and the source grammar is a significant subset of Python's own grammar. Programs and metalanguage code can be interleaved within the same file, so Ace is {\em language-integrated}. Ace programs are statically-typed, and types themselves are objects in the metalanguage. They can be equipped by developers with methods that control how operators associated with them are typechecked and compiled. Because of the relationship between this mechanism and the idea of {\em active libraries} \cite{activelibraries}, discussed further in Section \ref{related}, we call them {\em active types}, and thus Ace is {\em actively-typed}.
\subsection{Example 1: Hello, World!}
To demonstrate the somewhat unconventional structure of Ace programs and libraries, we begin in Listing \ref{hello} with a variant of the canonical ``Hello, World!'' example written using the \verb|OpenCL| module. We note that this module, which we use throughout the paper in examples, is simply a library like any other. The core of Ace gives no special treatment to it; it is distributed with Ace for convenience. Aspects of its implementation will be described in Section \ref{att}.

@@

\section{Ace for Researchers}
Thus far, we have been largely discussing the OpenCL module in our examples. However, Ace gives no preferential treatment to this module; it is implemented entirely using the user-facing mechanisms described in this section. A C99 module has also been developed (which, due to its similarity to OpenCL, shares many of its implementation details), but we do not discuss it further in this paper. Extensions complementing these are described in Section C below.

>>>>

In this paper, we introduce Ace, a programming language targeting professional end-users as well as researchers across high-performance computing and related domains. Ace has followed a principled design methodology guided by this more comprehensive set of design criteria in order to avoid many of the issues that have hindered previous language-based approaches in high-performance computing. These criteria, this design methodology, and the novel mechanisms and designs developed to satisfy these criteria constitute the generalizable contributions of this paper. We also hope that Ace itself will be useful to the HPC community\footnote{\small Ace is openly available at http://acelang.org/.}, and we describe several use cases and an initial case study in order to preliminarily validate it's utility.

@@

The examples in Section \ref{usage} are based on an internalization of the OpenCL kernel programming language as a library. This library is built atop a novel extensibility mechanism that we call \emph{active typechecking and translation} (AT\&T) and detail in Section \ref{att}. This mechanism relies on the core idea of representing types as metalanguage objects. We refer to such types as \emph{active types} by relation to \emph{active libraries} \cite{activelibraries} (see Section \ref{related}). Unlike prior approaches to extensibility that take a language-global view, which hinders interoperability, the AT\&T mechanism guarantees that extensions are composable by construction by limiting their scope to expressions having a type defined by that extension.

%Despite its simplicity and safety, this mechanism is flexible enough to define the entirety of the OpenCL kernel language (which is derived directly from C99) and, combined with Ace's type propagation scheme, substantially enhance it's ease-of-use. 
To demonstrate the flexibility of this mechanism beyond OpenCL, we continue in Section \ref{usecases} by outlining examples of other, higher-level parallel abstractions that can be cleanly implemented using AT\&T, including global address spaces, message passing and functional data parallelism. First-class support for each of these has required a new language in the past. AT\&T allows compile-time logic to be safely included within libraries, and thus these first-class abstractions can be  used naturally within a single program or workflow.

Ace can be used as a standalone language via the \verb|acec| compiler and also as an interactive compilation environment. This mode of use, described in Section \ref{compenv}, uses just-in-time specialization to  integrate the widely-adopted \verb|numpy| library (including its internal type system) and OpenCL's host API (by a mechanism also available to other host APIs, such as CUDA's) with the Ace compiler itself to enable the direct invocation of Ace functions from within Python scripts with minimal overhead. Achieving this deep level of integration makes use of Ace's representation of types as well.

To demonstrate the utility of Ace for scientific workloads, we describe in Section \ref{casestudy} a case study where Ace was used in this mode to develop a scientific simulation framework. This framework has been used to  specify and efficiently execute thousands of realizations of a large stochastic neural circuit model on clusters of GPUs, achieving the same performance as raw OpenCL code while being more modular and concise than was feasible before.

We conclude in Sections \ref{related} and \ref{discussion} with related work and a discussion of the limitations of Ace at the time of writing, as well as a discussion of planned future work further validating Ace and building upon the concept of active typing.

@@

\section{Language Design and Usage}\label{usage}
\subsection{High-Level Design}
Static type systems are powerful tools for programming language design and implementation. By tracking the type of a value statically, a typechecker can verify the absence of many kinds of errors over all inputs. This simplifies and increases the performance of the run-time system, as errors need not be detected dynamically using tag checks and other kinds of assertions. Many parallel programming abstractions are defined in terms of, or benefit from, a type system that enforces a communication protocol, ensures the consistency of data and simplifies the dynamics of the run-time system (see Section \ref{usecases} for examples). Because \textbf{verifiability} and \textbf{performance} are key criteria and static typing is a core technique, Ace is fundamentally statically-typed.

It is legitimate to ask, however, why dynamically-typed languages are so widely-used in HPC. One reason has to do with the criteria of \textbf{ease-of-use}. Statically-typed languages, particularly those that HPC programmers are exposed to, can be verbose, requiring explicit type annotations on every function and variable declaration,  separate header files, and other declarations that exist to assist the compiler. Cordy identifies the principle of \emph{conciseness} with elimination of
redundancy and the availability of reasonable defaults \cite{cordy1992hints}. The high-level languages used in HPC have, consistent with this principle, avoided most of this overhead. Ace was originally conceived to explore the question: does this require run-time mechanisms, or can one express a statically-typed, low-level language at least as powerful as C without the overhead that is typically associated with it?

Rather than designing a new syntax, or modifying the syntax of C, we chose to directly use, without modification, an existing language, Python's, syntax. This was not an arbitrary choice but rather a key mechanism by which Ace achieves \textbf{continuity}. Python is amongst the most widely-adopted languages in the world, and enjoys similar  ubiquity within HPC specifically \cite{pythonhpc}. By directly adopting Python's syntax, Ace's syntax is immediately familiar and acceptable to a significant portion of it's intended audience. Moreover, any tools that handle Python source code, including parsers, editors, style checkers and documentation generators, can be used without modification for handling Ace code.

\subsection{Example 1: Hello, World!}
\begin{codelisting}
\lstinputlisting{hello.py}
\caption{\texttt{[hello.py]} A basic Ace program demonstrating the two-phase structure of Ace programs and libraries.}
\label{hello}
\end{codelisting}
\begin{codelisting}
\begin{lstlisting}[style=Bash]
$ `acec hello.py`
Hello, compile-time world!
Goodbye, compile-time world!
$ `cat hello.cl`
__kernel void main() {
    printf("Hello, run-time world.");
}\end{lstlisting}
\caption{Compiling \texttt{hello.py} using the \texttt{acec} compiler.}
\label{helloout}
\end{codelisting}

To demonstrate the somewhat unconventional structure of Ace programs and libraries, we begin in Listing \ref{hello} with a variant of the canonical ``Hello, World!'' example written using the \verb|OpenCL| module. We note that this module, which we use throughout the paper in examples, is simply a library like any other. The core of Ace gives no special treatment to it; it is distributed with Ace for convenience. Aspects of its implementation will be described in Section \ref{att}.

@@

\section{Structure and Usage}
%Ace is an {\em actively-typed, language-integrated compilation environment}. Let us begin by defining each of these terms. 
%A {\em compilation environment} is a system that allows developers to programmatically control aspects of compilation directly. The language used to do so is called the {\em metalanguage} while the grammar that programs themselves are written in is called the {\em source grammar}. The metalanguage of Ace is Python and the source grammar is a significant subset of Python's own grammar. Programs and metalanguage code can be interleaved within the same file, so Ace is {\em language-integrated}. Ace programs are statically-typed, and types themselves are objects in the metalanguage. They can be equipped by developers with methods that control how operators associated with them are typechecked and compiled. Because of the relationship between this mechanism and the idea of {\em active libraries} \cite{activelibraries}, discussed further in Section \ref{related}, we call them {\em active types}, and thus Ace is {\em actively-typed}.
\section{Ace for Researchers}
Thus far, we have been largely discussing the OpenCL module in our examples. However, Ace gives no preferential treatment to this module; it is implemented entirely using the user-facing mechanisms described in this section. A C99 module has also been developed (which, due to its similarity to OpenCL, shares many of its implementation details), but we do not discuss it further in this paper. Extensions complementing these are described in Section C below.
